Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2022-09-07 paper/paper-infer.tex
save

Most traits and diseases are heritable, if not all. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Knowing how much a phenotype is heritable and polygenic can teach a lot about it.
Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.

@@

Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS (without needing an independent test set).
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
We call this new preferred set of 1,444,196 SNPs HapMap3+.

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.6\textwidth]{overview}}
	\caption{[OVERVIEW FIGURE OF WHAT LDPRED2 CAN PROVIDE?] \label{fig:overview}}
\end{figure}

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}). Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$, and because these estimates are higher than values reported in the literature.

@@

We then investigate using the extended set of HapMap3 variants proposed here (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.7-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1-8.2]).
A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502-0.514] instead of 0.324 [0.320-0.329]), which is also reflected in a larger predictive performance ($r^2$ in test set of 0.516 [0.508-0.524] instead of 0.344 [0.335-0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).

Finally, we investigate different transformations to apply to some continuous phenotypes used here.
Indeed, 49 of such phenotypes seems log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference from LDpred2-auto.
Note that we now use the HapMap3+ set of variants here.
We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation, hinting that it probably makes sense to transform these phenotypes.
We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
Except for bilirubin and lipoprotein(a) concentration, generally higher $h^2$ estimates and predictive performance $r^2$ are obtained with the RIN-transformation than the log-transformation.

@@

In this section, we use the new HapMap3+ set of variants constructed here. 
We recall that we have defined 431 independent LD blocks for this set (Methods).
We compute a per-block (local) $h^2$ estimate, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by one block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).

@@

Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals as before, and one from a large GWAS meta-analysis of 1.6M European individuals \cite[]{yengo2022saturated}.
Using the method developed in \cite{prive2021using}, we first infer the ancestry proportions of individuals included in the meta-analysis as 81.9\% from N.W. Europe, 9.5\% from E. Europe, 6.5\% from Finland, 1.5\% of Ashkenazi ancestry, 0.3\% from S.W. Europe, and 0.2\% from W. Africa. 
We can therefore use the same N.W. European LD with these GWAS summary statistics. 
Note that we use the HapMap3+ set of 1,444,196 SNPs here, but that we use only 1,013,499 SNPs for the GWAS meta-analysis (those overlapping and passing quality control).

Intercepts from LD Score regression are increasing with sample size: 1.01 (SE: 0.007) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
SNP heritability estimates are of 65.5\% (SE: 2.8), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively. And of 60.2\% [95\% CI: 57.4, 63.1], 62.9\% [61.7, 64.0], and 54.1\% [53.8, 54.4] with LDpred2-auto, respectively.
The reduced heritability estimation from the meta-analysis can be partly explained by three reasons: a reduced number of overlapping SNPs, more heterogeneity in the individuals from the meta-analysis compared to within the UK Biobank only, and less effect of assortative mating (which is strong in the UK Biobank due to the presence of many spouses, and probably contributes to overestimating the heritability of height there).
As expected, predictive performance $r^2$ are increasing with sample size, with 30.0\% [29.2, 30.8], 42.7\% [42.2, 43.1], and 46.9\% [46.8, 47.1], respectively.
Note that the first two estimates (from UKBB) are slightly overestimated, and that the last one is based on a smaller number of overlapping SNPs.
We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% versus 54.1\%).
Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.1\% [1.0, 1.4], 2.2\% [2.0, 2.4], and 5.7\% [5.4, 6.0], consistent with results of simulations with a large polygenicity (p=10\%). Therefore, we estimate that height has at least 50,000 causal variants.
We also identify 1744 SNPs with a probability of being causal larger than 95\%.
As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.72 [-0.76, -0.68], -0.74 [-0.76, -0.71], and -0.78 [-0.81, -0.75], respectively.
Finally, we compute per-annotation heritability estimates, to investigate functional enrichment. We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).

@@

[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]

>>>>

Most traits and diseases are heritable, if not all. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Knowing how much a phenotype is heritable and how much it is polygenic can teach a lot about it.
Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.

@@

Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
We call this new preferred set of 1,444,196 SNPs HapMap3+.

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.6\textwidth]{overview}}
	\caption{[TODO: ADD AN OVERVIEW FIGURE OF WHAT LDPRED2 CAN PROVIDE?] \label{fig:overview}}
\end{figure}

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}). Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature.

@@

We then investigate using the extended set of HapMap3 variants proposed here (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.

Finally, we investigate different transformations to apply to some continuous phenotypes used here.
Indeed, 49 of the phenotypes used here seem log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference from LDpred2-auto.
Note that we now use the HapMap3+ set of variants here.
We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation, hinting that it probably makes sense to transform these phenotypes.
We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
Except for bilirubin and lipoprotein(a) concentration, generally higher $h^2$ estimates and predictive performance $r^2$ are obtained with the RIN-transformation than the log-transformation.

@@

In this section, we use the extended set of variants constructed here, HapMap3+. 
We recall that we have defined 431 independent LD blocks for this set (Methods).
We compute per-block (local) $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by one block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).

@@

Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M European individuals \cite[]{yengo2022saturated}.
Using the method developed in \cite{prive2021using}, we first infer the ancestry proportions of individuals included in the meta-analysis as 81.9\% from N.W. Europe, 9.5\% from E. Europe, 6.5\% from Finland, 1.5\% of Ashkenazi ancestry, 0.3\% from S.W. Europe, and 0.2\% from W. Africa. 
We can therefore use the same N.W. European LD with these GWAS summary statistics. 
Note that we use the HapMap3+ set of 1,444,196 SNPs here, but that we use only 1,013,499 SNPs for the GWAS meta-analysis (those overlapping and passing quality control).

Intercepts from LD Score regression are increasing with sample size: 1.01 (SE: 0.007) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
SNP heritability estimates are of 65.5\% (SE: 2.8), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively. And of 60.2\% [95\% CI: 57.4, 63.1], 62.9\% [61.7, 64.0], and 54.1\% [53.8, 54.4] with LDpred2-auto, respectively.
The reduced heritability estimation from the meta-analysis can be partly explained by three reasons: a reduced number of overlapping SNPs, more heterogeneity in the individuals from the meta-analysis compared to within the UK Biobank only, and less effect of assortative mating (which is strong in the UK Biobank due to the presence of many spouses, and probably contributes to overestimating the heritability of height there).
As expected, predictive performance $r^2$ are increasing with sample size, with 30.0\% [29.2, 30.8], 42.7\% [42.2, 43.1], and 46.9\% [46.8, 47.1], respectively.
Note that the first two estimates (from UKBB) are slightly overestimated, and that the last one is based on a smaller number of overlapping SNPs.
We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% versus 54.1\%).
Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.1\% [1.0, 1.4], 2.2\% [2.0, 2.4], and 5.7\% [5.4, 6.0], consistent with results of simulations with a large polygenicity (p=10\%). Therefore, we estimate that height has at least 50,000 causal variants.
We also identify 1744 SNPs with a probability of being causal larger than 95\%.
As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.72 [-0.76, -0.68], -0.74 [-0.76, -0.71], and -0.78 [-0.81, -0.75], respectively.
Finally, we compute per-annotation heritability estimates to investigate functional enrichment. We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).

@@

\noindent[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]

\noindent[TODO: EXPORT LDREF FOR NEW HAPMAP3+]
